---
title: "phenology_modelling_ex1"
author: "Leandra Stengele"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
```


Error in libgit2::git_signature_default : 
  config value 'user.name' was not found
--> Resolved with: usethis::use_git_config(user.name = "lell-ski", user.email = "196680295+lell-ski@users.noreply.github.com")


## Get started

```{r}
install.packages("phenocamr")

# I will use the phenocamr package which 
# interfaces with the phenocam network API
# to download time series of vegetation 
# greenness and derived phenology metrics
library(phenocamr)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(GenSA)
library(terra)
#library(zoo)
```


```{r}
# download greenness time series,
# calculate phenology (phenophases),
# amend with DAYMET data
phenocamr::download_phenocam(
  site = "harvard$",
  veg_type = "DB",
  roi_id = "1000",
  daymet = TRUE,
  phenophase = TRUE,
  trim = 2022,
  out_dir = tempdir()
)

harvard_phenocam_data <- readr::read_csv(
  file.path(tempdir(), "harvard_DB_1000_3day.csv"), 
  comment = "#"
)

# reading in harvard phenology only retaining
# spring (rising) phenology for the GCC 90th
# percentile time series (the default)
harvard_phenology <- readr::read_csv(
  file.path(
    tempdir(),
    "harvard_DB_1000_3day_transition_dates.csv"
    ),
  comment = "#"
) |>
  dplyr::filter(
    direction == "rising",
    gcc_value == "gcc_90"
  )
```
Comment on Code: The :: in "phenocamr::download_phenocam" specifies where the "download_phenocam" comes from. (idk about this example but) this is very important in terms of shadowing. When theres e.g. the Fn Filter() which is contained in the packages dplyr and Stats and i have loaded both packages, it is unclear from which package i wanna take the filter() fn. therefore i do dplyr::filter().



Plot GCC
```{r}
ggplot(harvard_phenocam_data) +
  geom_line(
    aes(
      as.Date(date),
      smooth_gcc_90
    ),
    colour = "grey25"
  ) +
  geom_point(
    data = harvard_phenology,
    aes(
      as.Date(transition_25),
      threshold_25
    )
  ) +
  labs(
    x = "",
    y = "GCC"
  ) +
  theme_bw() +
  theme(
    legend.position = "none"
  )
```


implement GDD Fn (cumulative sum of temperatures above Threshhold T0 = 5°C)
```{r}
# return mean daily temperature as well
# as formal dates (for plotting)
harvard_temp <- harvard_phenocam_data |>
  group_by(year) |>
  dplyr::mutate(
    tmean = (tmax..deg.c. + tmin..deg.c.)/2
  ) |> 
  dplyr::mutate(
    date = as.Date(date),
    gdd = cumsum(ifelse(tmean >= 5, tmean - 5, 0))
  ) |>
  dplyr::select(
    date,
    year,
    tmean,
    gdd
  ) |>
  ungroup()

# convert the harvard phenology data and only
# retain required data
harvard_phenology <- harvard_phenology |>
  mutate(
    doy = as.numeric(format(as.Date(transition_25),"%j")),
    year = as.numeric(format(as.Date(transition_25),"%Y"))
  ) |>
  select(
    year,
    doy,
    transition_25,
    threshold_25
  )
```
Comment on Code: |> (former %>%) is the pipeline command that defines the order by which commands e.g. on a dataset should be executed. e.g. first filter the dataset, then isolate some values, then sort them etc. It saves the steps of doing the steps individually and after each step creating a new dataset with the new, saved changes (data1, data2, data3, ...).


plot the GDD
```{r}
# grab only the 2010 value of spring phenology
harvard_phenology_2010 <- harvard_phenology |>
  dplyr::filter(
    year == 2010
  )

harvard_gdd_value <- harvard_temp |>
  dplyr::filter(
    date == harvard_phenology_2010$transition_25
  )

p <- ggplot(harvard_temp) +
  geom_line(
    aes(
      date,
      tmean
    )
  ) +
  geom_point(
    aes(
      date,
      tmean,
      colour = tmean > 5,
      group = 1
    )
  ) +
  geom_vline(
    data = harvard_phenology_2010,
    aes(
      xintercept = as.Date(transition_25)
    )
    ) +
  scale_colour_discrete(
    type = c(
      "blue",
      "red"
      )
  ) +
  labs(
    x = "",
    y = "Temperature (deg. C)"
  ) +
  xlim(
    c(
    as.Date("2010-01-01"),
    as.Date("2010-06-30")
    )
  ) +
  theme_bw() +
  theme(
    legend.position = "none"
  )

p2 <- ggplot(harvard_temp) +
  geom_line(
    aes(
      date,
      gdd
    )
  ) +
  geom_point(
    aes(
      date,
      gdd,
      colour = tmean > 5,
      group = 1
    )
  ) +
  scale_colour_discrete(
    type = c(
      "blue",
      "red"
      )
  ) +
  geom_vline(
    data = harvard_phenology_2010,
    aes(
      xintercept = as.Date(transition_25)
      )
    ) +
  geom_hline(
    data = harvard_gdd_value,
    aes(
      yintercept = gdd
    ),
    lty = 2
    ) +
  labs(
    x = "",
    y = "GDD (deg. C)"
  ) +
  xlim(
    c(
    as.Date("2010-01-01"),
    as.Date("2010-06-30")
    )
  ) +
  ylim(c(0, 1000)) +
  theme_bw()  +
  theme(
    legend.position = "none"
  )

# compositing
library(patchwork)
p + p2 + 
  plot_layout(ncol = 1) + 
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")"
    )
```

# 6.3 Growing degree day model optimization
Our GDD-based leaf-out model can be written in the form of a function that takes the temperature time series as its first argument, and as two parameters the temperature threshold above which temperatures are accumulated, and the critical GDD that determines the DOY at which leaf-out is predicted.
```{r}
gdd_model <- function(temp, par) {
  #print(length(temp))
  # split out parameters from a simple
  # vector of parameter values
  temp_threshold <- par[1]
  gdd_crit <- par[2]
  # weights
  weight_factor <- par[3]
  day_idx <- c(1:length(temp))
  print(head(day_idx))
  specific_weight <- c(weight_factor*day_idx)
  print(head(specific_weight))
  
  
  
  # accumulate growing degree days for
  # temperature data
  gdd <- cumsum((ifelse(temp > temp_threshold, temp - temp_threshold, 0))*specific_weight)
  print((gdd))
  
  # figure out when the number of growing
  # degree days exceeds the minimum value
  # required for leaf development, only
  # return the first value
  doy <- unlist(which(gdd >= gdd_crit)[1])
  
  return(doy)
}


```

Running the model on the original 2010 data with the previously observed values and parameters 5C for the temperature threshold and 130.44C for the critical GDD, should yield a predicted leaf-out date that matches the observed leaf-out date of 114.
```{r}
# confirm that the model function
# returns expected results (i.e. DOY 114)
# (we filter out the year 2010, but
# removing the filter would run the
# model for all years!)
prediction <- harvard_temp |>
  dplyr::filter(
    year == 2010
  ) |>
  group_by(year) |>
  summarize(
    pred = gdd_model(
      temp = tmean,
      par = c(5, 130.44, 0.01)
    )  
  )
print(prediction)
```
## 6.3.1 Phenology model calibration
To obtain accurate predictions across a larger set of years and sites, we estimate a general set of parameters for our growing degree day model, i.e., we calibrate the model parameters. To do so, we want to minimize the error (a cost function) between the model results (for a given set of parameters, i.e. the temperature threshold and critical GDD) and our observed data obtained for multiple sites and years. 
```{r}
# run model and compare to true values
# returns the RMSE
rmse_gdd <- function(par, data) {
  
  # split out data
  drivers <- data$drivers
  validation <- data$validation
  
  # calculate phenology predictions
  # and put in a data frame
  predictions <- drivers |>
    group_by(year) |>
    summarise(
      predictions = gdd_model(
        temp = tmean,
        par = par
      )
    )
  
  predictions <- left_join(predictions, validation, by = "year")
  
  rmse <- predictions |>
    summarise(
      rmse = sqrt(mean((predictions - doy)^2, na.rm = TRUE))
    ) |>
    pull(rmse)
  
  # return rmse value
  return(rmse)
}
```

Defining the limits of the parameter space well can significantly reduce the time needed to converge on a solution
```{r}
# starting model parameters
par = c(0, 130, 0)

# limits to the parameter space
lower <- c(-10,0,0)
upper <- c(45,500,1)

# data needs to be provided in a consistent
# single data file, a nested data structure
# will therefore accept non standard data formats
data <- list(
  drivers = harvard_temp,
  validation = harvard_phenology
  )

# optimize the model parameters
optim_par = GenSA::GenSA(
 par = par,
 fn = rmse_gdd,
 lower = lower,
 upper = upper,
 control = list(
   max.call = 4000
   ),
 data = data
)$par
```

```{r}
# run the model for all years
# to get the phenology predictions
predictions <- harvard_temp |>
  group_by(year) |>
  summarize(
   prediction = gdd_model(
    temp = tmean,
    par = optim_par
  )  
  )
```

When looking at this small dataset for Harvard forest, we see a reasonable agreement between observed and predicted values (Figure 6.3). Obviously, including more sites and years would increase the model performance.
```{r}
# join predicted with observed data
validation <- left_join(predictions, harvard_phenology)

ggplot(validation) +
  geom_smooth(
    aes(
      doy,
      prediction
    ),
    colour = "grey25",
    method = "lm"
  ) +
  geom_point(
    aes(
      doy,
      prediction
    )
  ) +
  geom_abline(
    intercept=0, 
    slope=1, 
    linetype="dotted"
    ) +
  labs(
    x = "Observed leaf-out date (DOY)",
    y = "Predicted leaf-out date (DOY)"
  ) +
  theme_bw()  +
  theme(
    legend.position = "none"
  )
```
Figure 6.3: Scatterplot of observed and predicted leaf-out dates as day-of-year (DOY). A linear regression is overplot as a full dark grey line with confidence intervals shown as a light grey envelope. 


# 6.4 Spatial scaling
As example we will use a region around Boston in the eastern United States, defined by the coordinates -72 to -70 East, and 42 to 44 North. First I download both minimum and maximum temperature data and average them to a mean daily value (as previously used), using the appeears R package (Hufkens 2023).

In case there are issues with creating a NASA user account or the download you can use:
```{r}
local_file <- tempfile(fileext = ".nc")
download.file("https://github.com/fabern/handfull_of_pixels/raw/refs/heads/main/data/DAYMET.004_2012/DAYMET.004_1km_aid0001.nc",
              destfile = local_file)
r1 <- terra::rast(local_file)
# Assign the correct CRS, e.g. WGS84 (EPSG:4326)
terra::crs(r1) <- "epsg:4326"
```

```{r}
# Calculate the daily mean values based on 'tmin' and 'tmax'
mean_layer <- terra::mean(r1["tmax"], 
                          r1["tmin"])
# fix the variable naming
varnames(mean_layer) <- "tmean"                             
names(mean_layer) <- gsub("tmax","tmean",names(mean_layer))
```


```{r}
terra::plot(mean_layer)
```
Figure 6.4: Daily mean temperature for the first 16 days of 2012 for the greater Boston area. 


For convenience I limit the data to the first 180 days (layers) of the dataset to reduce the memory footprint of the calculations.
```{r}
# subset to first 180 days
ma_nh_temp <- terra::subset(
  mean_layer,
  1:180
)
```

One can then apply the model to this raster (cube) using the the terra::app() function and an appropriately formulated function, i.e. our growing degree day model gdd_model().
```{r}
predicted_phenology <- terra::app(
  ma_nh_temp,
  fun = gdd_model,
  par = optim_par
)
```

Plotting these results show a diverse response across the landscape, with higher elevation locations in the north west of the scene having a later leaf out date (in DOY) then coastal or urban areas (i.e. the greater Boston areas south east in the scene). This is both a reflection of the lapse rate as discussed in Chapter 4 and the urban heat island effect (Zhang, Friedl, and Schaaf 2004).

```{r}
library(leaflet)

# set te colour scale manually
pal <- colorNumeric(
  "magma",
  values(predicted_phenology),
  na.color = "transparent"
  )

# build the leaflet map
# using ESRI tile servers
# and the loaded demo raster
leaflet() |> 
  addProviderTiles(providers$Esri.WorldImagery, group = "World Imagery") |>
  addProviderTiles(providers$Esri.WorldTopoMap, group = "World Topo") |>
  addRasterImage(
    predicted_phenology,
    colors = pal,
    opacity = 0.8,
    group = "Phenology model results"
    ) |>
  addLayersControl(
    baseGroups = c("World Imagery","World Topo"),
    position = "topleft",
    options = layersControlOptions(collapsed = FALSE),
    overlayGroups = c("Phenology model results")
    ) |>
  addLegend(
    pal = pal,
    values = values(predicted_phenology),
    title = "DOY")
```
Figure 6.5: Interactive map of the spatially scaled optimized growing degree model using DAYMET daily mean temperature data for tile 11935, including the greater Boston area in the south-east to the White Mountains in the north-west. 


